---
title: "Practice: Point Estimation & Data Visualization"
subtitle: "Tuesday, June 17, 2025"
author: "your name here"
execute:
  echo: true
  warning: false
  message: false
format: 
  html:
    df-print: paged
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
editor: source
---

```{r}
#| echo: false
library(tidyverse)
library(gsheet)
mlp_data <- gsheet2tbl("https://docs.google.com/spreadsheets/d/1fgkPVXXmif9sXP1Gg7BEeRfKO3tayn1vvb2ep7eQqjs/edit?gid=0#gid=0")
```

## Summaries: Continuous Variables

Let's use `mean_median()` to summarize the continuous variables in the MLP dataset.

```{r}
mlp_data %>% 
  mean_median(friendship, tail_shimmer, magical_energy)
```

Let's use `mean_median()` to summarize the continuous variables in the MLP dataset by pony type.

```{r}
mlp_data %>% 
  group_by(type) %>%
  mean_median(friendship, tail_shimmer, magical_energy)
```

## Summaries: Categorical Variables

Let's use `n_pct()` to summarize the types of ponies in the dataset.

```{r}
mlp_data %>% 
  n_pct(type, rows = 4)
```

Let's use `n_pct()` to describe friendship index by type.

```{r}
mlp_data %>% 
  n_pct(friendship, type, rows = 4)
```

## Graphs: Box Plots

Construct a box plot for the tail shimmer of the ponies (*tail_shimmer*).

<center>
```{r}
mlp_data %>% ggplot(aes(x = tail_shimmer)) +
  geom_boxplot() +
  labs(x = "Tail Shimmer") +
  theme_bw() +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank())
```
</center>

## Graphs: Histograms

Construct a histogram for the flying speed of ponies (*flying_speed*).

<center>
```{r}
mlp_data %>% ggplot(aes(x = flying_speed)) +
  geom_histogram(bins = 15, 
                 color = "#2E7D32", 
                 fill = "#4CAF50") +
  labs(x = "Flying Speed", 
       y = "Number of Ponies") +
  theme_bw() 
```
</center>


Construct a histogram for the magical energy of ponies (*magical_energy*).

<center>
```{r}
mlp_data %>% ggplot(aes(x = magical_energy)) +
  geom_histogram(bins = 15, 
                 color = "#8B6C42", 
                 fill = "#F0E9DD") +
  labs(x = "Magical Energy",
       y = "Number of Ponies") +
  theme_bw() 
```

## Graphs: Bar Graphs

Construct a bar graph for the combined age and sex of ponies.

<center>
```{r}
mlp_data %>%
  count(sex) %>%
  ggplot(aes(x = sex, y = n)) +
  geom_col() +
  labs(x = "Age and Sex of Pony",
       y = "Number of Ponies")+
  theme_bw()
```
</center>

Construct a bar graph for the type of pony.

<center>
```{r}
mlp_data %>%
  count(type) %>%
  ggplot(aes(x = type, y = n)) +
  geom_col() +
  labs(x = "Type of Pony",
       y = "Number of Ponies")+
  theme_bw()
```

## Graphs: Scatterplots

Construct a scatterplot with magical energy (*magical_energy*) on the *x*-axis and tail shimmer (*tail_shimmer*) on the *y*-axis.

<center>
```{r}
mlp_data %>% ggplot(aes(y = tail_shimmer, x = magical_energy)) +
  geom_point(size = 2) +
  labs(x = "Magical Energy",
       y = "Tail Shimmer") +
  theme_bw()
```
</center>

Construct a scatterplot with magical energy (*magical_energy*) on the *x*-axis and flying speed (*flying_speed*) on the *y*-axis.

<center>
```{r}
mlp_data %>% ggplot(aes(x = flying_speed, y = tail_shimmer)) +
  geom_point(size = 2) +
  labs(y = "Tail Shimmer",
       x = "Flying Speed (km/h)") +
  theme_bw()
```
</center>
