---
title: "*t*-Test Assumptions<br>Wilcoxon Rank Sum<br>Wilcoxon Signed Rank"
subtitle: "June 26, 2025<br>Thursday"
execute:
  echo: true
  warning: false
  message: false
  error: true
format: 
  revealjs:
    theme: default
    css: mlp2.css
    self-contained: true
    slide-number: true
    footer: "STA4173 - Biostatistics - Summer 2025"
    width: 1600
    height: 900
    df-print: paged
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
editor: source
---

## Introduction: Topics

- Assumptions on *t*-tests:
    - One-sample
        - Normality
    - Independent samples
        - Normality
        - Equal variance
    - Dependent samples
        - Normality
    
- Nonparametric alternatives
    - Independent medians ($M_1-M_2$)
    - Dependent medians ($M_d$)
    
```{r}
#| echo: false
# remove when ssstats updated

independent_qq <- function(data, continuous, grouping) {
  # Capture variables using tidy evaluation
  variable_q <- rlang::enquo(continuous)
  group_q    <- rlang::enquo(grouping)
  
  # Extract column names as characters for later labeling
  variable_name <- rlang::as_name(variable_q)
  group_name    <- rlang::as_name(group_q)
  
  # Filter out missing group values and get group levels
  levels_group <- data %>%
    dplyr::filter(!is.na(!!group_q)) %>%
    dplyr::pull(!!group_q) %>%
    unique()
  
  if (length(levels_group) != 2) {
    stop(paste("Grouping variable must have exactly 2 levels. Found:", paste(levels_group, collapse = ", ")))
  }
  
  # Create QQ plots
  plots <- lapply(levels_group, function(level_val) {
    ggplot2::ggplot(dplyr::filter(data, !!group_q == level_val),
                    ggplot2::aes(sample = !!variable_q)) +
      ggplot2::stat_qq_line(linetype = "dashed", color = "black", linewidth = 1) +
      ggplot2::stat_qq(color = "#666666") +
      ggplot2::theme_bw() +
      ggplot2::labs(title = paste("QQ Plot for", variable_name, "|", group_name, "=", level_val),
                    x = "Theoretical Quantiles", y = "Sample Quantiles")
  })
  
  # Create histograms
  plots2 <- lapply(levels_group, function(level_val) {
    ggplot2::ggplot(dplyr::filter(data, !!group_q == level_val),
                    ggplot2::aes(x = !!variable_q)) +
      ggplot2::geom_histogram(color = "black", fill = "#666666", bins = 30) +
      ggplot2::theme_bw() +
      ggplot2::labs(title = paste("Histogram for", variable_name, "|", group_name, "=", level_val),
                    x = variable_name, y = "Count")
  })
  
  # Combine and arrange plots
  ggpubr::ggarrange(plotlist = c(plots, plots2))
}

dependent_qq <- function(data, col1, col2) {
  # Capture expressions
  var1_q <- rlang::enquo(col1)
  var2_q <- rlang::enquo(col2)
  
  # Get variable names as strings (for labeling)
  var1_name <- rlang::as_name(var1_q)
  var2_name <- rlang::as_name(var2_q)
  
  # Calculate differences
  diff_df <- data %>%
    dplyr::mutate(diff = !!var1_q - !!var2_q) %>%
    dplyr::select(diff) %>%
    dplyr::filter(!is.na(diff))
  
  # QQ Plot
  qq_plot <- ggplot2::ggplot(diff_df, ggplot2::aes(sample = diff)) +
    ggplot2::stat_qq_line(linetype = "dashed", color = "black", linewidth = 1) +
    ggplot2::stat_qq(color = "#666666") +
    ggplot2::theme_bw() +
    ggplot2::labs(
      title = paste("QQ Plot of Paired Differences:", var1_name, "-", var2_name),
      x = "Theoretical Quantiles",
      y = "Sample Quantiles"
    )
  
  # Histogram
  hist_plot <- ggplot2::ggplot(diff_df, ggplot2::aes(x = diff)) +
    ggplot2::geom_histogram(color = "black", fill = "#666666", bins = 30) +
    ggplot2::theme_bw() +
    ggplot2::labs(
      title = paste("Histogram of Paired Differences:", var1_name, "-", var2_name),
      x = "Difference",
      y = "Count"
    )
  
  # Arrange and return both plots
  ggpubr::ggarrange(qq_plot, hist_plot, ncol = 2)
}
```

## Introduction: Normality Assumption

- All *t*-tests assume approximate normality of the data.

    - In the case of one-sample *t*-tests, the measure of interest must somewhat follow a normal distribution.

    - In the case of two-sample *t*-tests, the measure of interest *in each group* must somewhat follow a normal distribution.
    
- Note that a paired *t*-test is technically a one-sample *t*-test, so we will examine normality of the difference. 

## Normality Assumption: Quantile-Quantile Plots

- There are formal tests for normality ([see article here](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3693611/)), however, we will not use them.
    - Tests for normality are not well-endorsed by statisticians.
    
- Instead, we will assess normality using a quantile-quantile (Q-Q) plot.

- A Q-Q plot helps us visually check if our data follows a specific distribution (here, the normal).
    - It compares the quantiles of our sample data to the quantiles of a theoretical distribution (the normal).

- How do we read Q-Q plots?
    - Each dot represents one observation in our dataset.
    - If the data follow a normal distribution, we will observe that the dots fall roughly along a straight diagonal line.
    - We focus on the "middle" of the graph.
    
## Normality Assumption: Quantile-Quantile Plots

<center>
```{r}
#| echo: false
library(ssstats)
library(tidyverse)
library(ggpubr)
set.seed(187782)
pegasi_flaps <- tibble(flaps = rnorm(100, mean = 120, sd = 10))
qq <- pegasi_flaps %>% ggplot(aes(sample = flaps)) +
  stat_qq_line(color = "gray", linewidth = 1) +
  stat_qq() +
  labs(x = "Theoretical Quantiles",
       y = "Sample Quantiles",
       title = "Quantile-Quantile Plot") +
  theme_bw()

hist <- pegasi_flaps %>% ggplot(aes(x = flaps)) +
  geom_histogram(fill = "gray", color = "black", bins = 10) +
  labs(title = "Histogram", x = "") + 
  theme_bw()

ggarrange(qq, hist, ncol = 2)
```
</center>

## Normality Assumption: Quantile-Quantile Plots

<center>
```{r}
#| echo: false
set.seed(881)
pegasi_flaps <- tibble(flaps = rexp(100, rate = 1/50))
qq <- pegasi_flaps %>% ggplot(aes(sample = flaps)) +
  stat_qq_line(color = "gray", linewidth = 1) +
  stat_qq() +
  labs(x = "Theoretical Quantiles",
       y = "Sample Quantiles",
       title = "Quantile-Quantile") +
  theme_bw()

hist <- pegasi_flaps %>% ggplot(aes(x = flaps)) +
  geom_histogram(fill = "gray", color = "black", bins = 10) +
  labs(title = "Histogram", x = "") + 
  theme_bw()

ggarrange(qq, hist, ncol = 2)
```
</center>

## Normality Assumption: Quantile-Quantile Plots

<center>
```{r}
#| echo: false
set.seed(188277)
pegasi_flaps <- tibble(flaps = rt(100, df = 3) * 10 + 50)
qq <- pegasi_flaps %>% ggplot(aes(sample = flaps)) +
  stat_qq_line(color = "gray", linewidth = 1) +
  stat_qq() +
  labs(x = "Theoretical Quantiles",
       y = "Sample Quantiles",
       title = "Quantile-Quantile") +
  theme_bw()


hist <- pegasi_flaps %>% ggplot(aes(x = flaps)) +
  geom_histogram(fill = "gray", color = "black", bins = 10) +
  labs(title = "Histogram", x = "") + 
  theme_bw()

ggarrange(qq, hist, ncol = 2)
```
</center>

## Normality Assumption: Quantile-Quantile Plots

<center>
```{r}
#| echo: false
set.seed(658236)
pegasi_flaps <- tibble(flaps = rbeta(100, shape1 = 5, shape2 = 1) * 100)
qq <- pegasi_flaps %>% ggplot(aes(sample = flaps)) +
  stat_qq_line(color = "gray", linewidth = 1) +
  stat_qq() +
  labs(x = "Theoretical Quantiles",
       y = "Sample Quantiles",
       title = "Quantile-Quantile") +
  theme_bw()

hist <- pegasi_flaps %>% ggplot(aes(x = flaps)) +
  geom_histogram(fill = "gray", color = "black", bins = 10) +
  labs(title = "Histogram", x = "") + 
  theme_bw()

ggarrange(qq, hist, ncol = 2)
```
</center>
    
## Normality Assumption: Independent Means

```{r}
#| echo: false

set.seed(8651)
wing_flap <- tibble(pony_names = c("Rainbow Dash", "Soarin", "Spitfire", "Cloudchaser", "Fleetfoot", "Lightning Dust", "Windy Whistles", "Stormy Skies", "Gusty Gale", "Zephyr Breeze", "Skydancer", "Aero Ace", "Nimbus Sprint", "Breeze Wing", "Flash Fly", "Twister Tail", "Sirensong", "Aero Wave", "Sky Dancer", "Gale Gust", "Storm Strider", "Cyclone Swift", "Wind Whirl", "Jet Stream", "Flurry Flap"),
                    wing_flap_rate = round(rnorm(n = length(pony_names), mean = 50, sd = 8),1),
                    target = dplyr::if_else(wing_flap_rate>=50, "Above", "Below"),
                    apples = dplyr::if_else(target == "Above", round(rnorm(n = length(pony_names), mean = 25, sd = 2)), round(rnorm(n = length(pony_names), mean = 15, sd = 3))),
                    pre_training_wfr = round(rnorm(n = length(pony_names), mean = 50, sd = 8),1),
                    post_training_wfr =round(rnorm(n = length(pony_names), mean = 50, sd = 4),1))
```

- Recall our example from last lecture: In the skies above Cloudsdale, Pegasus trainers believe that an average healthy Pegasus flaps its wings 50 flaps per minute when cruising. To see if today's young Pegasi conform to that standard, a researcher samples 25 Pegasi at the Cloudsdale Training Grounds and measures each pony's wing‐flap rate (in flaps/minute). <br>

```{r}
#| echo: false
set.seed(9671)
wing_flap %>% select(-pre_training_wfr, -post_training_wfr) %>% slice_sample(n=5)
```

## Normality Assumption: Independent Means

- Further, we performed a two-sample *t*-test to determine if the above target pegasi are eating 5 or more apples than the below target pegasi ($\alpha=0.05$). 

```{r}
wing_flap %>% independent_mean_HT(grouping = target,
                                  continuous = apples, 
                                  mu = 5, 
                                  alternative = "greater", 
                                  alpha = 0.05)
```

- Are our results valid?

## Normality Assumption: Independent Means (R)

- We will use the `independent_qq()` function from `library(ssstats)` to assess normality.

```{r}
#| eval: false
dataset_name %>% independent_qq(continuous = continuous_variable,
                                grouping = grouping_variable)
```

- This will provide the the Q-Q plots and the histograms for the two independent groups under consideration.

## Normality Assumption: Independent Means

- Let's now look at the normality assumption for our example.

- How should we change the code for our dataset?

```{r}
#| eval: false
dataset_name %>% independent_qq(continuous = continuous_variable,
                                grouping = grouping_variable)
```

## Normality Assumption: Independent Means

- Let's now look at the normality assumption for our example.

- How should we change the code for our dataset?

```{r}
#| eval: false
wing_flap %>% independent_qq(continuous = apples,
                             grouping = target)
```

## Normality Assumption: Independent Means

- Running the code,

<center>
```{r}
wing_flap %>% independent_qq(continuous = apples,
                             grouping = target)
```
</center>

## Introduction: Variance Assumption

- In addition to normality, the two-sample *t*-test assumes equal variance between groups.
    - *Homoskedastic*: same variance
    - *Heteroskedastic*: different variances

- We can check this assumption and easily adjust if the assumption is broken.

- Graphical method: scatterplot of residuals

- Formal method: test for equal variances (Brown-Forsythe-Levine)

## Variance Assumption: Residual Plot (R)

- We will use the `plot_residuals()` function from `library(ssstats)` to graphically assess the assumption of equal variance.

```{r}
#| eval: false
dataset_name %>% plot_residuals(continuous = continuous_variable,
                                grouping = grouping_variable)
```

- We should evaluate and compare the lengths of the resulting "lines".

## Variance Assumption: Residual Plot

- Let's now look at the normality assumption for our example.

- How should we change the code for our dataset?

```{r}
#| eval: false
dataset_name %>% plot_residuals(continuous = continuous_variable,
                                grouping = grouping_variable)
```

## Variance Assumption: Residual Plot

- Let's now look at the normality assumption for our example.

- Our updated code:

<center>
```{r}
#| eval: false
wing_flap %>% plot_residuals(continuous = apples,
                             grouping = target)
```
</center>

## Variance Assumption: Residual Plot

- Running the code,

<center>
```{r}
#| eval: false
wing_flap %>% plot_residuals(continuous = apples,
                             grouping = target)
```
</center>

## Hypothesis Testing: Two or More Variances

- If we believe the assumption may be violated, we can test for equal variance using the Brown-Forsythe-Levine (BFL) test.

- This test is valid for more than two groups (read: we will see it again!)

- **Hypotheses**
    - $H_0: \ \sigma^2_1 = \sigma^2_2 = ... = \sigma^2_k$
    - $H_1:$ at least one $\sigma^2_i$ is different.
    
## Hypothesis Testing: Two or More Variances

- **Test Statistic:**

$$
F_0 = \frac{\sum_{i=1}^k n_i(\bar{z}_{i.}-\bar{z}_{..})^2/(k-1)}{\sum_{i=1}^k \sum_{j=1}^{n_i} (z_{ij}-\bar{z}_{i.})^2/(N-k)} \sim F_{\text{df}_{\text{num}}, \text{df}_{\text{den}}}
$$

- where
    - $n_i$ is the sample size of group $i$
    - $\bar{z}_{i.}$ is the median of group $i$ and $\bar{z}_{..}$ is the grand median
    - $k$ is the number of groups
    - $N = \sum_{i=1}^k n_i$ is the overall sample size
    - $\text{df}_{\text{num}} = k-1$, $\text{df}_{\text{den}} = (N-k)$.

## Hypothesis Testing: Two or More Variances

- Note that the BFL is a one-tailed test, which is different than when we are testing means using the $t$ distribution.

- **$p$-value:**

$$p = P\left[F_{\text{df}_{\text{num}}, \text{df}_{\text{den}}} \ge F_0\right]$$

## Hypothesis Testing: Two or More Variances (R)

```{r}
wing_flap %>% variances_HT(continuous = apples,
                           grouping = target)

```

## Variance Assumption: Independent Means 

- example set up

## Variance Assumption: Independent Means

- example set up with correct code

## Variance Assumption: Independent Means 

- code has been run

## Variance Assumption: Independent Means

- HT results

## Variance Assumption: Broken -- Now What?

$$ \text{df}=\frac{ \left( \frac{s^2_1}{n_1} + \frac{s_2^2}{n_2} \right)^2 }{ \frac{(s_1^2/n_1)^2}{n_1-1} + \frac{(s_2^2/n_2)^2}{n_2-1} }$$


## Normality Assumption: Dependent Means

- normality assumption for dependent t

## Normality Assumption: Dependent Means (R)

- syntax for qq

## Normality Assumption: Dependent Means

- example set up

## Normality Assumption: Dependent Means

- example set up with correct code

## Normality Assumption: Dependent Means

- code has been run

## Wrap Up: *t*-Test Assumptions

- **Important note!!** 
    - *I do not expect you to agree with my assessment of q-q plots!* 
    - What I do expect is that you know what to do after making your assessment.
        - Meet assumption $\to$ use *t*-test
        - Do not meet assumption $\to$ use nonparametric equivalent
        
- Next: nonparametrics!        

## Introduction: Nonparametrics

- The *t*-tests we have already learned are considered *parametric* methods.

    - There is a distributional assumption on the test.
    
- *Nonparametric* methods do not have distributional assumptions.

    - We typically transform the data to their ranks and then perform calculations.

- Why don't we always use nonparametric methods?

    - They are often less efficient: a larger sample size is required to achieve the same probability of a Type I error.
    
    - They discard useful information :(
    
## Introduction: Wilcoxon Rank Sum

- introduce W RS
    
## Wilcoxon Rank Sum: Ranking Data

- In the nonparametric tests we will be learning, the data will be ranked.

- Let us first consider a simple example, $$x: \ 1, 7, 10, 2, 6, 8$$

- Our first step is to reorder the data: $$x: \ 1, 2, 6, 7, 8, 10$$

- Then, we replace with the ranks: $$R: \ 1, 2, 3, 4, 5, 6$$

## Wilcoxon Rank Sum: Ranking Data

- What if all data values are not unique? We will assign the <u>average</u> rank for that group.

- For example, $$x: \ 9, 8, 8, 0, 3, 4, 4, 8$$

- Let's reorder:$$x: \ 0, 3, 4, 4, 8, 8, 8, 9$$

- Rank ignoring ties:$$R: \ 1, 2, 3, 4, 5, 6, 7, 8$$

- Now, the final rank:$$R: \ 1, 2, 3.5, 3.5, 6, 6, 6, 8$$

## Wilcoxon Rank Sum{.smaller}

**Hypotheses**

- $H_0:$ $M_1 - M_2 = M_0$ | $H_0:$ $M_1 - M_2 \le M_0$ | $H_0:$ $M_1 - M_2 \ge M_0$
- $H_1:$ $M_1 - M_2 \ne M_0$ | $H_1:$ $M_1 - M_2 > M_0$ | $H_1:$ $M_1 - M_2 < M_0$

**Test Statistic & *p*-Value**

- $T = \sum R_{\text{sample 1}} - \frac{n_1(n_1+1)}{2}$
- $p =$ (calculated by R :))

**Rejection Region**

- Reject $H_0$ if $p < \alpha$.

**Conclusion/Interpretation**

- [Reject or fail to reject] $H_0$.

- There [is or is not] sufficient evidence to suggest [alternative hypothesis in words].

## Wilcoxon Rank Sum (R)

- syntax

## Wilcoxon Rank Sum: Independent Medians

- example set up

## Wilcoxon Rank Sum: Independent Medians

- example with correct code

## Wilcoxon Rank Sum: Independent Medians

- code has been run

## Wilcoxon Rank Sum: Independent Medians

- HT typeset

# Introduction: Wilcoxon Signed Rank

## Wilcoxon Signed Rank: Ranking Data

- Before ranking, we will find the difference between the paired observations and eliminate any 0 differences.

    - Note 1: elimniating 0 differences is the big difference between the other tests! 
    
    - Note 2: because we are eliminating 0 differences, this means that our sample size will update to the number of pairs with a non-0 difference.
    
- When ranking, we the differences are ranked based on the absolute value of the difference.

- We also keep the sign of the difference.

    - We will have positive ranks and negative ranks.

| *X* | *Y* | *D* | \|*D*\| | Rank |
|-----|-----|-----|---------|------|
| 5   | 8   | -3  | 3       | $-$ 1.5 |
| 8   | 5   | 3   | 3       | $+$ 1.5  |
| 4   | 4   | 0   | 0       | ---------   |

## Wilcoxon Signed Rank{.smaller}

**Hypotheses**

- $H_0:$ $M_d = M_0$ | $H_0:$ $M_d \le M_0$ | $H_0:$ $M_d \ge M_0$
- $H_1:$ $M_d \ne M_0$ | $H_1:$ $M_d > M_0$ | $H_1:$ $M_d < M_0$

**Test Statistic & *p*-Value**

- $T_0 = \min(T+,|T_-|)$ if two-tailed, $T_0 = T_+$ if left-tailed, and $T_0 = |T_-|$ if right-tailed.
- $p =$ (calculated by R :))

**Rejection Region**

- Reject $H_0$ if $p < \alpha$.

**Conclusion/Interpretation**

- [Reject or fail to reject] $H_0$.

- There [is or is not] sufficient evidence to suggest [alternative hypothesis in words].

## Wilcoxon Signed Rank (R)

- syntax

## Wilcoxon Signed Rank: Dependent Medians

- example set up

## Wilcoxon Signed Rank: Dependent Medians

- example with correct code

## Wilcoxon Signed Rank: Dependent Medians

- code has been run

## Wilcoxon Signed Rank: Dependent Medians

- HT typeset

## Putting It All Together

- When asked to compare two groups, I first must decide:
    - Independent data?
    - Dependent data?
    
- Then, I must decide:
    - Do I meet the assumptions for the appropriate *t*-test?
        - If so $\to$ proceed with *t*-test.
        - If not $\to$ use nonparametric alternative.
        
- Remember the pairings!
    - Independent *t*-test $\to$ Wilcoxon rank sum.
    - Dependent *t*-test $\to$ Wilcoxon signed rank.
    
## "Use the Appropriate Test": Independent or Dependent?

- Example 1 - independent or dependent?

## "Use the Appropriate Test": *t* or Nonparametric?

- Example 1 - parametric or non?

## "Use the Appropriate Test": Independent or Dependent?

- Example 2 - independent or dependent?

## "Use the Appropriate Test": *t* or Nonparametric?

- Example 2 - parametric or non?

## "Use the Appropriate Test": Independent or Dependent?

- Example 3 - independent or dependent?

## "Use the Appropriate Test": *t* or Nonparametric?

- Example 3 - parametric or non?

## "Use the Appropriate Test": Independent or Dependent?

- Example 4 - independent or dependent?

## "Use the Appropriate Test": *t* or Nonparametric?

- Example 4 - parametric or non?

## Wrap Up        

- Today's lecture:
    - Normality assumption (all *t*) $\to$ Q-Q plot
    - Variance assumption (independent *t*) $\to$ folded *F* test
    - Wilcoxon rank sum (nonparametric equivalent to independent *t*)
    - Wilcoxon signed rank (nonparametric equivalent to dependent *t*)
    
- Next class:   
    - Project 1!

## Wrap Up    

- Daily activity: .qmd is available on Canvas.
    - Due date: Monday, June 30, 2025.

- You will upload the resulting .html file on Canvas. 
    - Please refer to the help guide on the Biostat website if you need help with submission.
    
- Housekeeping:
    - Are you in the Discord server?
    - Do you have questions for me?
    - Do you need my help with anything from prior lectures?
