---
title: "Two-Way ANOVA"
subtitle: "July 15, 2025<br>Tuesday"
execute:
  echo: true
  warning: false
  message: false
  error: true
format: 
  revealjs:
    theme: default
    css: mlp2.css
    self-contained: true
    slide-number: true
    footer: "STA4173 - Biostatistics - Summer 2025"
    width: 1600
    height: 900
    df-print: paged
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
editor: source
---

## Introduction: Two-Way ANOVA

- Recall that ANOVA allows us to compare the means of three or more groups.

- In one-way ANOVA, we are only considering one factor (grouping variable). 

- Now, we will discuss two-way ANOVA, which allows us to consider a second factor (grouping variable). 

- We now partition the SS<sub>Trt</sub> into the different factors under consideration.
    
- Recall that SS<sub>E</sub> is the "catch all" for unexplained variance. 

    - When we add factors to our model, we are moving part of the SS<sub>E</sub> into the SS<sub>Trt</sub>.
    
## Introduction: Two-Way ANOVA

- Let's discuss some of the language used in two-way ANOVA. 

- Factor A has $a$ levels.
    
    - Pony type: earth, pegasus, unicorn

- Factor B has $b$ levels.

    - Preferred apple: honeycrisp, granny

- There are $ab$ treatment groups.

## Introduction: Two-Way ANOVA

- Now that we are including two factors, we must consider the interaction term. 

    - The relationship between [outcome] and [factor 1] depends on the level of [factor 2].

- In our example, suppose Pinkie Pie is testing a new apple pie recipe and has ponies taste then rate the new recipe,

    - the relationship between rating and preferred apple depends on the pony type.
    - the relationship between rating and pony type depends on the preferred apple. 

## Introduction: Two-Way ANOVA

- The ANOVA table that we are working to construct: 

| **Source** | **Sum of Squares** | **df** | **Mean Square** | ***F*** |
|-|-|-|-|-|
A | SS<sub>A</sub> | df<sub>A</sub> | MS<sub>A</sub> | *F*<sub>A</sub>
B | SS<sub>B</sub> | df<sub>B</sub> | MS<sub>B</sub> | *F*<sub>B</sub>
AB | SS<sub>AB</sub> | df<sub>AB</sub> | MS<sub>AB</sub> | *F*<sub>AB</sub>
Error | SS<sub>E</sub> | df<sub>E</sub> | MS<sub>E</sub> |
Total | SS<sub>Tot</sub> | df<sub>Tot</sub> | | |   

## Two-Way ANOVA: Computation

- Let there be $a$ levels of factor A and $b$ levels of factor B. 

    - $y_{ijk}$ is the observation on the $k^{th}$ experimental unit receiving the $i^{th}$ level of factor A and the $j^{th}$ level of factor B 
    - $y_{i.}$ is the sum for all observations at the $i^{th}$ level of factor A,
    - $y_{.j}$ is the sum for all observations at the $j^{th}$ level of factor B,
    - $y_{ij}$ is the sum for observations at the $i^{th}$ level of factor A and the $j^{th}$ level of factor B,
    - $y_{..}$ is the sum of all observations, 
    - $(y^2)_{..}$ is the sum of the squared observations, 
    - $n$ is the number in each group (of $a \times b$ treatments)

## Two-Way ANOVA: Computation

- To find the SS and df:
$$
\begin{align*}
\text{SS}_{\text{A}} &= \frac{\sum_i y_{i.}^2}{bn}  - \frac{(y_{..})^2}{abn} & \text{df}_{\text{A}} &= a-1 \\
\text{SS}_{\text{B}} &= \frac{\sum_j y_{.j}^2}{an} - \frac{(y_{..})^2}{abn} & \text{df}_{\text{B}} &= b-1 \\
\text{SS}_{\text{AB}}&= \frac{\sum_{ij} y_{ij}^2}{n} - \frac{(y_{..})^2}{abn} - \text{SS}_{\text{A}} - \text{SS}_{\text{B}} & \text{df}_{\text{AB}} &= (a-1)(b-1) \\
\text{SS}_{\text{E}} &= \text{SS}_{\text{Tot}} - \text{SS}_{\text{A}} - \text{SS}_{\text{B}} - \text{SS}_{\text{AB}} & \text{df}_{\text{E}} &= ab(n-1)  \\
\text{SS}_{\text{Tot}} &= (y^2)_{..} - \frac{(y_{..})^2}{abn} & \text{df}_{\text{Tot}} &= abn-1
\end{align*}  
$$  

## Two-Way ANOVA: Computation

- To find the MS:

$$\text{MS}_{\text{X}} = \frac{\text{SS}_{\text{X}}}{\text{df}_{\text{X}}}$$

- To find the test statistic:

$$\text{F}_{\text{X}} = \frac{\text{MS}_{\text{X}}}{\text{MS}_{\text{E}}}$$

## Two-Way ANOVA: ANOVA Table (R)

- We will use the `two_way_ANOVA_table()` function from `library(ssstats)` to construct the two-way ANOVA table.

```{r}
#| eval: false
dataset_name %>% two_way_ANOVA_table(continuous = continuous_variable,
                                     A = factor_A,
                                     B = factor_B)
```

## Two-Way ANOVA: Example Set Up

```{r}
#| echo: false
library(tidyverse)
library(ssstats)
set.seed(421)                 # for reproducibility
n_per_cell <- 35              # 35 students of each type in each setting

# Define means that create an interaction
means <- tribble(
  ~type,      ~setting,     ~mu,
  "Earth",    "Library",     78,
  "Earth",    "Cafe",        79,
  "Pegasus",  "Library",     75,
  "Pegasus",  "Cafe",        85,   # big boost in café
  "Unicorn",  "Library",     88,   # big boost in library
  "Unicorn",  "Cafe",        80
)

sigma <- 5  # common SD

# Generate the dataset
grades <- means %>% 
  uncount(n_per_cell, .id = "student") %>%      # expand rows
  mutate(
    grade = rnorm(n(), mean = mu, sd = sigma)
  )
```

- At Friendship University, college-aged ponies enroll in the introductory STEM course "Applied Equestrian Engineering" (data is collected in **grades**).  Researchers want to know how two factors influence the overall grade (percentage scale):
    - Pony Type (*type*; different innate talents and study habits): earth, pegasus, unicorn
    - Study Setting (*setting*; varied levels of background stimulation): Quiet Library or Sugarcube Corner Cafe
    
```{r}
#| echo: false
grades %>% group_by(type) %>% mean_median(grade)
```

## Two-Way ANOVA: Example Set Up

- At Friendship University, college-aged ponies enroll in the introductory STEM course "Applied Equestrian Engineering"(data is collected in **grades**).  Researchers want to know how two factors influence the overall grade (percentage scale):
    - Pony Type (*type*; different innate talents and study habits): earth, pegasus, unicorn
    - Study Setting (*setting*; varied levels of background stimulation): Quiet Library or Sugarcube Corner Cafe
    
```{r}
#| echo: false
grades %>% group_by(setting) %>% mean_median(grade)
```

## Two-Way ANOVA: Example Set Up

- At Friendship University, college-aged ponies enroll in the introductory STEM course "Applied Equestrian Engineering."  Researchers want to know how two factors influence the overall grade (percentage scale), pony type and study setting.
    
```{r}
#| echo: false
grades %>% group_by(type, setting) %>% mean_median(grade)
```

## Two-Way ANOVA: ANOVA Table

- At Friendship University, college-aged ponies enroll in the introductory STEM course "Applied Equestrian Engineering" (data is collected in **grades**).  Researchers want to know how two factors influence the overall grade (percentage scale):
    - Pony Type (*type*; different innate talents and study habits): earth, pegasus, unicorn
    - Study Setting (*setting*; varied levels of background stimulation): Quiet Library or Sugarcube Corner Cafe

- Let's construct the two-way ANOVA table. How should we change the following code?

```{r}
#| eval: false
dataset_name %>% two_way_ANOVA_table(continuous = continuous_variable,
                                     A = factor_A,
                                     B = factor_B)
```

## Two-Way ANOVA: ANOVA Table

- At Friendship University, college-aged ponies enroll in the introductory STEM course "Applied Equestrian Engineering" (data is collected in **grades**).  Researchers want to know how two factors influence the overall grade (percentage scale):
    - Pony Type (*type*; different innate talents and study habits): earth, pegasus, unicorn
    - Study Setting (*setting*; varied levels of background stimulation): Quiet Library or Sugarcube Corner Cafe

- Let's construct the two-way ANOVA table. Our updated code:

```{r}
#| eval: false
grades %>% two_way_ANOVA_table(continuous = grade,
                               A = type,
                               B = setting)
```

## Two-Way ANOVA: ANOVA Table

- Running the code:

```{r}
#| echo: false
grades %>% two_way_ANOVA_table(continuous = grade,
                               A = type,
                               B = setting)
```

## Two-Way ANOVA: Testing Interactions

- **Hypotheses**

    - $H_0:$ there is not an interaction between [factor A] and [factor B]
    - $H_1:$ there is an interaction between [factor A] and [factor B]

- **Test Statistic and *p*-Value**

    - $F_{\text{AB}}$ 
    - $p = P[F_{\text{df}_{\text{AB}}, \text{df}_{\text{E}}} \ge F_{\text{AB}}]$

- **Rejection Region**

    - Reject $H_0$ if $p<\alpha$.
    
## Two-Way ANOVA: Testing Interactions (R)

- We will use the `two_way_ANOVA_HT()` function from `library(ssstats)` to perform the test for the interaction.

```{r}
#| eval: false
dataset_name %>% two_way_ANOVA_HT(continuous = continuous_variable,
                                  A = factor_A,
                                  B = factor_B,
                                  interaction = TRUE,
                                  alpha = specified_alpha)
```
    
## Two-Way ANOVA: Testing Interactions

- At Friendship University, college-aged ponies enroll in the introductory STEM course "Applied Equestrian Engineering" (data is collected in **grades**).  Researchers want to know how two factors influence the overall grade (percentage scale):
    - Pony Type (*type*; different innate talents and study habits): earth, pegasus, unicorn
    - Study Setting (*setting*; varied levels of background stimulation): Quiet Library or Sugarcube Corner Cafe

- Determine if there is an interaction between pony type and study setting. Test at the $\alpha=0.01$ level. 

- How should we update the following code?    
    
```{r}
#| eval: false
dataset_name %>% two_way_ANOVA_HT(continuous = continuous_variable,
                                  A = factor_A,
                                  B = factor_B,
                                  interaction = TRUE,
                                  alpha = specified_alpha)
```  
    
## Two-Way ANOVA: Testing Interactions

- At Friendship University, college-aged ponies enroll in the introductory STEM course "Applied Equestrian Engineering" (data is collected in **grades**).  Researchers want to know how two factors influence the overall grade (percentage scale):
    - Pony Type (*type*; different innate talents and study habits): earth, pegasus, unicorn
    - Study Setting (*setting*; varied levels of background stimulation): Quiet Library or Sugarcube Corner Cafe

- Determine if there is an interaction between pony type and study setting. Test at the $\alpha=0.01$ level. 

- Our updated code,
    
```{r}
#| eval: false
grades %>% two_way_ANOVA_HT(continuous = grade,
                            A = type,
                            B = setting,
                            interaction = TRUE,
                            alpha = 0.01)
```

## Two-Way ANOVA: Testing Interactions

- Running the code,
    
```{r}
grades %>% two_way_ANOVA(continuous = grade,
                         A = type,
                         B = setting,
                         interaction = TRUE,
                         alpha = 0.01)
```
  
## Two-Way ANOVA: Testing Interactions

- **Hypotheses**

    - $H_0:$ there is not an interaction between pony type and study setting
    - $H_1:$ there is an interaction between pony type and study setting

- **Test Statistic and *p*-Value**

    - $F_{\text{AB}} = 52.45$; $p < 0.001$

- **Rejection Region**

    - Reject $H_0$ if $p<\alpha$; $\alpha=0.01$.
    
- **Conclusion/Interpretation**

    - Reject $H_0$. There is sufficient evidence to suggest that the relationship between grades and study setting depends on pony type.

## Two-Way ANOVA: Profile Plots

- What happens after testing for an interaction? 
    - If significant (reject $H_0$), we can construct a profile plot to visualize what's going on and help explain the effect. 
    - If not significant (FTR $H_0$)... stay tuned.
    
- Profile plot: a plot of treatment group means.
    - *y*-axis: always the average outcome
    - *x*-axis: either factor A or B
    - lines on the plot: the factor that was not selected for the *x*-axis

- Note that this is just a graph of the means! It's valid to construct a profile plot even if the interaction is not sigificant.

## Two-Way ANOVA: Profile Plots (R)

- We will use the `profile_plot()` function from `library(ssstats)` to construct basic profile plots.

```{r}
#| eval: false
dataset_name %>% profile_plot(continuous = continuous_variable,
                              xaxis = variable_on_x,
                              lines = variable_for_lines)
```

## Two-Way ANOVA: Profile Plots

- At Friendship University, college-aged ponies enroll in the introductory STEM course "Applied Equestrian Engineering" (data is collected in **grades**).  Researchers want to know how two factors influence the overall grade (percentage scale).

- Let's now construct the profile plot with pony type (*type*) on the *x*-axis and create the lines using study setting (*setting*).

- How should we update this code?

```{r}
#| eval: false
dataset_name %>% profile_plot(continuous = continuous_variable,
                              xaxis = variable_on_x,
                              lines = variable_for_lines)
```

## Two-Way ANOVA: Profile Plots

- At Friendship University, college-aged ponies enroll in the introductory STEM course "Applied Equestrian Engineering" (data is collected in **grades**).  Researchers want to know how two factors influence the overall grade (percentage scale).

- Let's now construct the profile plot with pony type (*type*) on the *x*-axis and create the lines using study setting (*setting*).

- Our updated code,

```{r}
#| eval: false
grades %>% profile_plot(continuous = grade,
                        xaxis = type,
                        lines = setting)
```

## Two-Way ANOVA: Profile Plots

- Running the code,

<center>
```{r}
grades %>% profile_plot(continuous = grade,
                        xaxis = type,
                        lines = setting)
```
</center>

## Two-Way ANOVA: Profile Plots

- What happens if we switch *xaxis* and *lines*?

<center>
```{r}
grades %>% profile_plot(continuous = grade,
                        xaxis = setting,
                        lines = type)
```

## Two-Way ANOVA: Testing Main Effects

- What if the interaction *is not* significant?
    - We want to *remove* the interaction and examine the *main effects*.
    - That is, we want to look at the individual factors.
    
- We will reconstruct the ANOVA table. If we were doing this by hand:
    1. Rewrite ANOVA table without interaction; do not change SS, df, and MS for the main effects and total.
    2. Update the error term: add the SS<sub>AB</sub> to SS<sub>E</sub> and df<sub>AB</sub> to df<sub>E</sub>.
    3. Recalculate MS<sub>E</sub>.
    4. Recalculate *F*<sub>A</sub> and *F*<sub>B</sub>, then find the corresponding *p*-values.
    
## Two-Way ANOVA: Testing Main Effects (R)
    
- We will use the `two_way_ANOVA()` function from `library(ssstats)` to perform the test for main effects.

```{r}
#| eval: false
dataset_name %>% two_way_ANOVA(continuous = continuous_variable,
                               A = factor_A,
                               B = factor_B,
                               interaction = FALSE,
                               alpha = specified_alpha)
```

## Two-Way ANOVA: Testing Main Effects

```{r}
#| echo: false
set.seed(1203)

n_per_cell <- 30
sigma <- 6

# Define means with additive structure (no interaction)
means <- tribble(
  ~club,         ~study,     ~mu,
  "Club",        "Solo",     85,
  "Club",        "Group",    88,  # +3 for group study
  "No Club",     "Solo",     80,  # -5 overall
  "No Club",     "Group",    83   # +3 for group study
)

# Simulate scores
quiz_scores <- means %>% 
  uncount(n_per_cell, .id = "student") %>% 
  mutate(
    score = rnorm(n(), mean = mu, sd = sigma)
  )
```

- At Ponyville High, students in a general science class (**quiz_scores**) were divided based on their membership in the Science Club (yes/no; *club*) and their primary method of studying outside of school (solo/group; *study*).

- Researchers want to know whether these factors impact science quiz scores (out of 100 points; *score*). 

- Let's first check for an interaction ($\alpha=0.05$). How should we change the following code?

```{r}
#| eval: false
dataset_name %>% two_way_ANOVA(continuous = continuous_variable,
                               A = factor_A,
                               B = factor_B,
                               interaction = TRUE or FALSE,
                               alpha = specified_alpha)
```

## Two-Way ANOVA: Testing Main Effects

- At Ponyville High, students in a general science class (**quiz_scores**) were divided based on their membership in the Science Club (yes/no; *club*) and their primary method of studying outside of school (solo/group; *study*).

- Researchers want to know whether these factors impact science quiz scores (out of 100 points; *score*). 

- Let's first check for an interaction ($\alpha=0.05$). Our updated code,

```{r}
#| eval: false
quiz_scores %>% two_way_ANOVA(continuous = score,
                              A = club,
                              B = study,
                              interaction = TRUE)
```

## Two-Way ANOVA: Testing Main Effects

- Running the code, 

```{r}
quiz_scores %>% two_way_ANOVA(continuous = score,
                              A = club,
                              B = study,
                              interaction = TRUE)
```

## Two-Way ANOVA: Testing Main Effects

- At Ponyville High, students in a general science class (**quiz_scores**) were divided based on their membership in the Science Club (yes/no; *club*) and their primary method of studying outside of school (solo/group; *study*).

- Researchers want to know whether these factors impact science quiz scores (out of 100 points; *score*). 

- Now let's remove the interaction. How should we update our code?

```{r}
#| eval: false
quiz_scores %>% two_way_ANOVA(continuous = score,
                              A = club,
                              B = study,
                              interaction = TRUE)
```

## Two-Way ANOVA: Testing Main Effects

- At Ponyville High, students in a general science class (**quiz_scores**) were divided based on their membership in the Science Club (yes/no; *club*) and their primary method of studying outside of school (solo/group; *study*).

- Researchers want to know whether these factors impact science quiz scores (out of 100 points; *score*). 

- Now let's remove the interaction. Our updated code,

```{r}
#| eval: false
quiz_scores %>% two_way_ANOVA(continuous = score,
                              A = club,
                              B = study,
                              interaction = FALSE)
```

## Two-Way ANOVA: Testing Main Effects

- **Hypotheses**

    - $H_0:$ $\mu_1$ = $\mu_2$ = ... = ($\mu_a$ or $\mu_b$)
    - $H_1:$ At least one $\mu_i$ is different.

- **Test Statistic and *p*-Value**

    - $F_{\text{X}}$ (pull from ANOVA table)
    - $p = P[F_{\text{df}_{\text{X}}, \text{df}_{\text{E}}} \ge F_{\text{X}}]$

## Two-Way ANOVA: Testing Main Effects

- Running the code,

```{r}
#| echo: false
quiz_scores %>% two_way_ANOVA(continuous = score,
                              A = club,
                              B = study,
                              interaction = FALSE)
```


## Two-Way ANOVA: Testing Main Effects

- **Hypotheses**
    - $H_0: \ \mu_{\text{club}} = \mu_{\text{no club}}$
    - $H_1: \ \mu_{\text{club}} \ne \mu_{\text{no club}}$

- **Test Statistic and *p*-Value**
    - $F_{\text{club}} = 11.19$; $p = 0.001$

- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$.
    
- **Conclusion/Interpretation**
    - Reject $H_0$. There is sufficient evidence to suggest that the quiz grades between those in the science club are different from those not in the science club.    
    
## Two-Way ANOVA: Testing Main Effects

- **Hypotheses**
    - $H_0: \ \mu_{\text{solo}} = \mu_{\text{group}}$
    - $H_1: \ \mu_{\text{solo}} \ne \mu_{\text{group}}$

- **Test Statistic and *p*-Value**
    - $F_{\text{study}} = 11.24$; $p = 0.001$

- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$.
    
- **Conclusion/Interpretation**
    - Reject $H_0$. There is sufficient evidence to suggest that the quiz grades between those in the science club are different from those not in the science club.       
    
## Two-Way ANOVA: Assumptions   

- The ANOVA assumptions we learned last week hold true.

- We assume that the error term follows a normal distribution with mean 0 and a constant variance, $\sigma^2$. i.e.,
$$\varepsilon_{ij} \overset{\text{iid}}{\sim} N(0, \sigma^2)$$

- Very important note: **the assumption is on the error term** and NOT on the outcome!

- We will use the residual (the difference between the observed value and the predicted value) to assess assumptions:

$$ e_{ij} = y_{ij} - \hat{y}_{ij} $$

## Two-Way ANOVA: Assumptions (R)

- We will use the `ANOVA2_assumptions()` function from `library(ssstats)` to request the graphs necessary to asssess our assumptions.

```{r}
#| eval: false
dataset_name %>% ANOVA2_assumptions(continuous = continuous_variable,
                                    A = factor_A,
                                    B = factor_B,
                                    interaction = TRUE or FALSE)
```

## Two-Way ANOVA: Assumptions

- At Ponyville High, students in a general science class (**quiz_scores**) were divided based on their membership in the Science Club (yes/no; *club*) and their primary method of studying outside of school (solo/group; *study*). Researchers want to know whether these factors impact science quiz scores (out of 100 points; *score*). 

- Let's now check the ANOVA assumptions. How should we edit the following code?

```{r}
#| eval: false
dataset_name %>% ANOVA2_assumptions(continuous = continuous_variable,
                                    A = factor_A,
                                    B = factor_B,
                                    interaction = TRUE or FALSE)
```

## Two-Way ANOVA: Assumptions

- At Ponyville High, students in a general science class (**quiz_scores**) were divided based on their membership in the Science Club (yes/no; *club*) and their primary method of studying outside of school (solo/group; *study*). Researchers want to know whether these factors impact science quiz scores (out of 100 points; *score*). 

- Let's now check the ANOVA assumptions. How should we edit the following code?

```{r}
#| eval: false
dataset_name %>% ANOVA2_assumptions(continuous = continuous_variable,
                                    A = factor_A,
                                    B = factor_B,
                                    interaction = TRUE or FALSE)
```

## Two-Way ANOVA: Assumptions

- At Ponyville High, students in a general science class (**quiz_scores**) were divided based on their membership in the Science Club (yes/no; *club*) and their primary method of studying outside of school (solo/group; *study*). Researchers want to know whether these factors impact science quiz scores (out of 100 points; *score*). 

- Let's now check the ANOVA assumptions. Our updated code,

```{r}
#| eval: false
quiz_scores %>% ANOVA2_assumptions(continuous = score,
                                  A = club,
                                  B = study,
                                  interaction = TRUE)
```

## Two-Way ANOVA: Assumptions

- Running the code,

<center>
```{r}
quiz_scores %>% ANOVA2_assumptions(continuous = score,
                                  A = club,
                                  B = study,
                                  interaction = TRUE)
```
</center>

## Wrap Up

- Today we have covered two-way ANOVA.
    - Two-way ANOVA table
    - Testing interactions
    - Testing main effects (only when interaction is not present)
    - Profile plot (plot of group means)
    - ANOVA assumptions
    
- Note about nonparametric two-way ANOVA.
    - Methodology is "newish" - we are not covering.
    
- Thursday: Assignment 2    

## Wrap Up    

- Daily activity: .qmd is available on Canvas.
    - Due date: Monday, July 21, 2025.

- You will upload the resulting .html file on Canvas. 
    - Please refer to the help guide on the Biostat website if you need help with submission.
    
- Housekeeping:
    - Quiz at 1:15!
    - Do you have questions for me?
    - Do you need my help with anything from prior lectures? Practices? Project 1?
